Turbo fast no compromise gravitational wave parameter estimation

### 2023-01-12 src/tex/ms.tex
Got most of the acknowledgement down

The main objective in PE can be summarized as given data in the form of a time
series of strain $\mathbf{d}$, find the distribution of the source parameters
$\mathbf{\theta}$ that is compatible with the data, i.e.
$p(\mathbf{\theta}|\mathbf{d})$. To compute this object, we can rewrite
it using Bayes' theorem,
\begin{align}
    p(\theta| d) = \frac{\mathcal{L}(d|\theta)\pi(\theta)}{p(d)},
\end{align}
where $\mathcal{L}(d|\theta)$ is the likelihood function, $\pi(\theta)$ is the
prior distribution, and $p(d)$ is the evidence. Since the evidence is a
normalization constant that does not depend on the source parameters, it is
often omitted if we are only interested in the posterior distribution. The prior
distribution is often chosen to be some simple distribution, such as uniformly
distributed in the component masses or a Gaussian distribution in the spins.
Assuming the noise follows from a stationary Gaussian process, the
log-likelihood in the case of GW is given by
\begin{align}
    \log{\mathcal{L}(d|\theta)} = -\left<d-h(\mathbf{\theta})|d-h(\mathbf{\theta})\right>/2,
\label{eq: loglikelihood}
\end{align}
where $d$ is the strain data, $h(\mathbf{\theta})$ is the strain predicted by a
waveform model with a specific set of source parameters. The right hand side of
eq. \ref{eq: loglikelihood} can be evaluated in either time or frequency domain.
Since it is computationally cheaper to compute the likelihood in the frequency
domain, we choose to compute the likelihood throughout this work, which can be defined as 
\te{Need to take the real part?}
\begin{align}
    % \left<a|b\right> = 4 Re\int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
    \left<a|b\right> = 4 \int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
\label{eq: innerproduct}
\end{align}
where $\mathcal{S}_n(f)$ is the noise power spectral density (PSD). To compute
the integral shown in eq. \ref{eq: innerproduct}, we need to evaluate a chosen
waveform model $h(\mathbf{\theta})$ at a number of frequnecy sample points.
Evaluating the likelihood function is often the most computationally intensive
part of the PE. The most accurate waveform model is numerical relativity (NR).
However, depending on the source parameters, generating one time series of
strain can take a day to half a year, which makes NR prohibitively
expensive for PE. To circumvent this problem, there are several waveform
"approximant" families, including the IMRPhenom family \kw{}, SEOB family \kw{},
and NR surrogate family\kw{cite}. For shorter events such as a $30-30\ M_{\odot}$
BBH, one waveform call could take $10\text{ms}$ to $\sim 1s$ \kw{verify number
in lalsuite}. For longer events such as a $1.4-1.4\ M_{\odot}$ BNS event, the
evaluation time could go up to \kw{Fill}. Since one needs to evaluate the
likelihood many times during sampling\footnote{A typical PE run with
\texttt{Bilby} takes $>10^6$ likelihood evaluation to converge.}, the
computational cost in evaluating the waveform accumulate and is the main reason
of the long runtime of GW PE.

@@

% Observation of traits of flowMC
This means independent chains can now jump between modes, or within a single
mode that exhibits complex geometry that cannot be handled only through the
local sampler, which greatly reduces both the burn-in time and sampling time.

While the method would still work when there is only one independent local
sampler chain, this method works much better when we employ a lot of independent
chains because of the increased number of samples available for training. 
\kw{Mention how this helps gravitational wave, gives examples.}

In practice, we find the normalizing flow not only helps jumping betw

@@

% Compared to other works
There are recent works from different groups on speeding up parameter estimation
of gravitational wave, including efficient reparameterization and deep learning
\kw{cite}. While all of these methods can get down to minutes-scale parameter
estimation with high fidelity, we would like to highlight the unique strength of
this work, and the potential interplay between our work and others' work.

% Stephen's work
Compared to \kw{Stephan's paper}, we do not rely on pre-training the neural
network on a large collection of waveforms and noise realization. This means as
soon as new waveform models and noise models are available, our algorithm can
already be deployed. Furthermore, our method is essentially an MCMC algorithm,
which inherits the merit of convergence measures in MCMC. As we are only using
the normalizing flow as a proposal distribution, and the normalizing flow is
trained jointly with a local sampler, we do not suffer from overfitting since
our training data is being generated on the fly and is always approaching the
target distribution. In this sense, we do not introduce potential extra
systematic error to inference results.

@@

%Tousif's work
Compared to \kw{Tousif's paper}, we do not rely on handcrafted
reparameterization of the coordinate systems. If the reparameterization scheme
is known ahead of time, a nice reparameterization is also encouraged. However,
handcrafted reparameterization depends on the assumptions used in deriving the
reparameterization scheme, which would inevitably run into limitations of use
cases. Our work can be viewed as an automatic reparameterization powered by
normalizing flow, while not being exact hence as efficient as an explicit
reparameterization, the method proposed in this work applies to a much
border class of problems, such as parameter estimation with precessing
waveforms, testing GR, and multi-event joint inference.

It is always beneficial to reparameterize if the reparameterization is known
ahead of time. For the class of problems where \kw{Tousif's paper} applies, we
can incorporate the reparameterization scheme into our MCMC pipeline to reduce
the complexity of the problem, hence speeding up the training phase. On the
other hand, if there is part of the target posterior that is not included in the
reparameterization scheme, the normalizing flow should still be able to learn to
produce accurate samples efficiently.

@@

% Additional traits for having differentiable waveforms
There are some future developments we are working on. While IMRPhenomD is
a reasonable start, it lacks some qualitative features that other
state-of-the-art models have, such as precession, higher mode, and eccentricity \kw{cite}. It
also has a higher mismatch with reference numerical relativity waveforms compared
to more recent waveform models. Currently, we are working on building differentiable
IMRPhenomPv2 and NRSurrogate waveforms. Going forward, we encourage the waveform
development community to leverage autodiff environments such as Jax when
constructing new waveform models. Having a differentiable waveform model is not
only beneficial for parameter estimation, but also for other applications such
as calibrating waveforms to numerical relativity results, as detailed in
\kw{cite}.

% Access to higher dimensional problem
While standard GW analysis goes up to 17 dimensions. non-standard GW PE problems
can have more parameters, which could potentially lead to more complicated
geometry in the target posterior that is hard to reparameterize. For example,
there are \kw{Be specific with the number} parameterized modifications to the
waveform one can introduce when testing GR, which means there are \kw{the number}
extra parameters one would need to consider in the PE process. Worse, these
parameters often introduce non-trivial degeneracies such as local correlation
and multi-modality. Currently testing GR is limited to varying these
modifications one at a time, partially due to the bottleneck in the sampler. Given
the gradient-based and normalizing flow-enhanced sampler, our code shows
the potential in solving this problem in full at once.

@@

To perform realistic PE with our pipeline, the antenna pattern of the
detector network needed to be taken into account. Our current code can perform
parameter estimation for any combination of the Hanford, Livingston, and Virgo
detectors, under the assumption of short signals such that the effect of Earth's
rotation can be ignored. For next-generation detector networks such as the
Einstein Telescope and the Cosmic Explorer \kw{cite}, a differentiable version
of their antenna pattern is needed.

% More features such as marginalization
The features we have implemented are the barebone version of parameter
estimation. We do not include marginalization schemes such as time, phase, and
calibration lines marginalization. Because of the reasonable performance of the sampler on
accelerators, time and phase marginalization is not necessary, as the
performance of our implementation is not significantly impacted by having two
extra dimensions. Other marginalization modes should be considered in the future.

% JIT overhead
Jax's JIT compilation drastically reduces the computational time to evaluate the
likelihood. However, it comes with a compilation overhead when the likelihood is
evaluated for the first time. We observe that the compilation time could depend
on the device where the code will be run. This is expected since Jax leverage
Accelerated Linear Algebra (XLA) takes advantage of accelerators, which means
Jax needs to compile the code for the specific device according to its
architecture. On an Nvidia A100 GPU, the compilation overhead could go up to 5
minutes for the waveform we are using. For the cases we have studied, the time
needed to obtain converging results on an A100 is about 2-3 minutes. This means
the compilation overhead is dominating the wall-clock time of the specific PE
run we considered. To utilize our implementation to its full potential, we are
looking into ways to reduce the compilation overhead or to cache the
compilation results to avoid paying the compilation overhead for every event.

@@

Another overhead we are looking into reducing is the time needed to find the
reference waveform for heterodyning the likelihood. Currently, we are using
\texttt{differential evolution} available in \texttt{scipy} to find the waveform
parameters which maximize the likelihood. Since \texttt{differnetial evolution}
has not been implemented in Jax, and the Jax waveform we use is not compatible
with the parallelization scheme in the \texttt{scipy} library, it takes around 5
minutes \kw{Benchmark it in the code and make this number precise.} to find the
reference waveform parameters for GW170817. There are two ways to improve the
performance in finding the parameters of the reference waveform. First, we can
explore different optimization strategy that is more compatible with the
strength of our pipeline, in particular differentiability of our likelihood and
the possibility to evaluate many waveforms in parallel with a GPU. Particle
swarm or stochastic gradient descent methods are promising candidates which we
intend to investigate in the future.

>>>>

The main objective in PE can be summarized as given data in the form of a time
series of strain $\mathbf{d}$, find the distribution of the source parameters
$\mathbf{\theta}$ that is compatible with the data, i.e.
$p(\mathbf{\theta}|\mathbf{d})$. To compute this object, we can rewrite
it using Bayes' theorem,
\begin{align}
    p(\theta| d) = \frac{\mathcal{L}(d|\theta)\pi(\theta)}{p(d)},
\end{align}
where $\mathcal{L}(d|\theta)$ is the likelihood function, $\pi(\theta)$ is the
prior distribution, and $p(d)$ is the evidence. Since the evidence is a
normalization constant that does not depend on the source parameters, it is
often omitted if we are only interested in the posterior distribution. The prior
distribution is often chosen to be some simple distribution, such as uniformly
distributed in the component masses or a Gaussian distribution in the spins.
Assuming the noise follows from a stationary Gaussian process, the
log-likelihood in the case of GW is given by
\begin{align}
    \log{\mathcal{L}(d|\theta)} = -\left<d-h(\mathbf{\theta})|d-h(\mathbf{\theta})\right>/2,
\label{eq: loglikelihood}
\end{align}
where $d$ is the strain data, $h(\mathbf{\theta})$ is the strain predicted by a
waveform model with a specific set of source parameters. The right hand side of
eq. \ref{eq: loglikelihood} can be evaluated in either time or frequency domain.
Since it is computationally cheaper to compute the likelihood in the frequency
domain, we choose to compute the likelihood throughout this work, which can be defined as 
\te{Need to take the real part?}
\begin{align}
    % \left<a|b\right> = 4 Re\int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
    \left<a|b\right> = 4 \int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
\label{eq: innerproduct}
\end{align}
where $\mathcal{S}_n(f)$ is the noise power spectral density (PSD). To compute
the integral shown in eq. \ref{eq: innerproduct}, we need to evaluate a chosen
waveform model $h(\mathbf{\theta})$ at a number of frequnecy sample points.
Evaluating the likelihood function is often the most computationally intensive
part of the PE. The most accurate waveform model is numerical relativity (NR).
However, depending on the source parameters, generating one time series of
strain can take a day to half a year, which makes NR prohibitively expensive for
PE. To circumvent this problem, there are several waveform "approximant"
families, including the IMRPhenom family \cite{Khan:2015jqa,
Garcia-Quiros:2020qpx}, SEOB family \cite{PhysRevD.89.061502}, and NR surrogate
family\cite{Varma:2019csw}. For shorter events such as a $30-30\ M_{\odot}$ BBH, one
waveform call could take $10\text{ms}$ to $\sim 1s$ \kw{verify number in
lalsuite}. For longer events such as a $1.4-1.4\ M_{\odot}$ BNS event, the
evaluation time could go up to \kw{Fill}. Since one needs to evaluate the
likelihood many times during sampling\footnote{A typical PE run with
\texttt{Bilby} takes $>10^6$ likelihood evaluation to converge.}, the
computational cost in evaluating the waveform accumulate and is the main reason
of the long runtime of GW PE.

@@

@@

% Compared to other works
There are recent works from different groups on speeding up parameter estimation
of gravitational wave, including efficient reparameterization
\cite{Islam:2022afg,Roulet:2022kot} and deep learning
\cite{Dax:2021tsq,Dax:2022pxd}. While all of these methods can get down to
minutes-scale parameter estimation with high fidelity, we would like to
highlight the unique strength of this work, and the potential interplay between
our work and others' work.

% Stephen's work
Compared to \cite{Dax:2021tsq,Dax:2022pxd}, we do not rely on pre-training the neural
network on a large collection of waveforms and noise realization. This means as
soon as new waveform models and noise models are available, our algorithm can
already be deployed. Furthermore, our method is essentially an MCMC algorithm,
which inherits the merit of convergence measures in MCMC. As we are only using
the normalizing flow as a proposal distribution, and the normalizing flow is
trained jointly with a local sampler, we do not suffer from overfitting since
our training data is being generated on the fly and is always approaching the
target distribution. In this sense, we do not introduce potential extra
systematic error to inference results.

@@

%Tousif's work
Compared to \cite{Islam:2022afg,Roulet:2022kot}, we do not rely on handcrafted
reparameterization of the coordinate systems. If the reparameterization scheme
is known ahead of time, a nice reparameterization is also encouraged. However,
handcrafted reparameterization depends on the assumptions used in deriving the
reparameterization scheme, which would inevitably run into limitations of use
cases. Our work can be viewed as an automatic reparameterization powered by
normalizing flow, while not being exact hence as efficient as an explicit
reparameterization, the method proposed in this work applies to a much border
class of problems, such as parameter estimation with precessing waveforms,
testing GR, and multi-event joint inference.

It is always beneficial to reparameterize if the reparameterization is known
ahead of time. For the class of problems where
\cite{Islam:2022afg,Roulet:2022kot} applies, we can incorporate the
reparameterization scheme into our MCMC pipeline to reduce the complexity of the
problem, hence speeding up the training phase. On the other hand, if there is
part of the target posterior that is not included in the reparameterization
scheme, the normalizing flow should still be able to learn to produce accurate
samples efficiently.

@@

% Additional traits for having differentiable waveforms
There are some future developments we are working on. While IMRPhenomD
\cite{Khan:2015jqa} is a reasonable start, it lacks some qualitative features
that other state-of-the-art models have, such as precession, higher mode, and
eccentricity. It also has a higher mismatch with reference numerical relativity
waveforms compared to more recent waveform models. Currently, we are working on
building differentiable IMRPhenomPv2 \cite{Khan:2018fmp} and three-parameters
NRSurrogate waveforms \cite{Varma:2018mmi}. Going forward, we encourage the
waveform development community to leverage autodiff environments such as Jax
when constructing new waveform models. Having a differentiable waveform model is
not only beneficial for parameter estimation, but also for other applications
such as calibrating waveforms to numerical relativity results, as detailed in
\kw{cite}.

% Access to higher dimensional problem
While standard GW analysis goes up to 17 dimensions. non-standard GW PE problems
can have more parameters, which could potentially lead to more complicated
geometry in the target posterior that is hard to reparameterize. For example,
\cite{LIGOScientific:2021sio} introduces 10 extra parameters to modify the
post-Newtonian coefficients predicted in GR, which should be taken care at the
same time during inference. On top of the increase of dimensionality, these
parameters often introduce non-trivial degeneracies such as local correlation
and multi-modality. Therefore, currently testing GR is limited to varying these
modifications one at a time, partially due to the bottleneck in the sampler.
Given the gradient-based and normalizing flow-enhanced sampler, our code shows
the potential in solving this problem in full at once.

@@

To perform realistic PE with our pipeline, the antenna pattern of the
detector network needed to be taken into account. Our current code can perform
parameter estimation for any combination of the Hanford, Livingston, and Virgo
detectors, under the assumption of short signals such that the effect of Earth's
rotation can be ignored. For next-generation detector networks such as the
Einstein Telescope and the Cosmic Explorer, a differentiable version
of their antenna pattern is needed.

% More features such as marginalization
The features we have implemented are the barebone version of parameter
estimation. We do not include marginalization schemes such as time, phase, and
calibration lines marginalization \cite{2019PASA...36...10T}. Because of the reasonable performance of the
sampler on accelerators, time and phase marginalization is not necessary, as the
performance of our implementation is not significantly impacted by having two
extra dimensions. Nevertheless, having less parameters in a PE is always
appreciated, which we are looking into including in our code as well as other
marginalization modes.

% JIT overhead
Jax's JIT compilation drastically reduces the computational time to evaluate the
likelihood. However, it comes with a compilation overhead when the likelihood is
evaluated for the first time. We observe that the compilation time could depend
on the device where the code will be run. This is expected since Jax leverage
Accelerated Linear Algebra (XLA) takes advantage of accelerators, which means
Jax needs to compile the code for the specific device according to its
architecture. On an Nvidia A100 GPU, the compilation overhead could go up to 8
minutes for the waveform we are using. For the cases we have studied, the time
needed to obtain converging results on an A100 is about 2-3 minutes. This means
the compilation overhead is dominating the wall-clock time of the specific PE
run we considered. To utilize our implementation to its full potential, we are
looking into ways to reduce the compilation overhead or to cache the
compilation results to avoid paying the compilation overhead for every event.

@@

Another overhead we are looking into reducing is the time needed to find the
reference waveform for heterodyning the likelihood. Currently, we are using
\texttt{differential evolution} \cite{Storn1997DifferentialE} available in
\texttt{scipy} to find the waveform parameters which maximize the likelihood.
Since \texttt{differnetial evolution} has not been implemented in Jax, and the
Jax waveform we use is not compatible with the parallelization scheme in the
\texttt{scipy} library, it takes around 5 minutes \kw{Benchmark it in the code
and make this number precise.} to find the reference waveform parameters for
GW170817. There are two ways to improve the performance in finding the
parameters of the reference waveform. First, we can explore different
optimization strategy that is more compatible with the strength of our pipeline,
in particular differentiability of our likelihood and the possibility to
evaluate many waveforms in parallel with a GPU. Particle swarm \cite{7869491}
and stochastic gradient descent methods \cite{10.5555/304710.304720} are
promising candidates which we intend to investigate in the future.
